The Spatial and Temporal Domains of Modern Ecology

### 2018-02-02 paper/main.tex
Edited ms and SI for near final submission

\author
{Lyndon Estes$^{\ast1, 2}$, Paul R. Elsen$^{3, 4}$, Tim Treuer$^{4}$, Labeeb Ahmed$^{5}$, \\
Kelly Caylor$^{6, 7}$, Jason Chang$^{5}$, Jonathan J. Choi$^{5}$, and Erle Ellis$^{6}$ \\
\\
\normalsize{$^{1}$Graduate School of Geography, Clark University, Worcester MA 01610, USA}\\
\normalsize{$^{2}$Woodrow Wilson School, Princeton University, Princeton, NJ 08544, USA}\\
\normalsize{$^{3}$Department of Environmental Science, Policy, and Management,}\\
\normalsize{University of California Berkeley, Berkeley, CA 94720, USA}\\
\normalsize{$^{4}$Ecology and Evolutionary Biology, Princeton University, Princeton, NJ 08544, USA}\\
\normalsize{$^{5}$Geography and Environmental Systems, University of Maryland Baltimore County,}\\ 
\normalsize{Baltimore, MD 21250, USA}\\
\normalsize{$^{6}$Department of Geography, University of California Santa Barbara, Santa Barbara, CA 93106, USA}\\
\normalsize{$^{7}$Bren School of Environmental Science and Management,}\\
\normalsize{University of California Santa Barbara, Santa Barbara, CA 93106, USA}\\
\normalsize{$^\ast$To whom correspondence should be addressed; E-mail:  LEstes@clarku.edu.}}
% Include the date command, but leave its argument blank.

@@

\section*{Abstract}
\textbf{To understand ecological phenomena, it is necessary to observe their behavior across multiple spatial and temporal scales. This need was first highlighted in the 1980s, and since then technology has opened previously inaccessible scales to observation. Have there been corresponding changes in the scales observed by modern ecologists? To help answer this question, we analyzed the resolution, extent, interval, and duration of observations (excluding experiments) in 348 studies published between 2004-2014. We found that observational scales were generally narrow because ecologists still primarily use conventional field techniques. In the spatial domain, most observations had resolutions $\leq$1 m$^2$ and extents $\leq$10,000 ha. In the temporal domain, most observations were either unreplicated or infrequently repeated ($\geq$1 month interval), and $\leq$1 year in duration. Compared to studies conducted prior to 2004, observational durations and resolutions appear largely unchanged, but intervals have become finer and extents larger. We also found a large gulf between the scales at which phenomena are actually observed, and the scales those observations ostensibly represent, raising concerns about observational comprehensiveness. Furthermore, most studies did not clearly report scale, suggesting that it remains a minor concern. We suggest that ecologists can obtain better understanding of the scales represented by observations by incorporating autocorrelation measures, while journals can promote attentiveness to scale by implementing scale reporting standards.}

%To fully understand ecological phenomena, we must observe them across multiple spatial and temporal scales. Beginning in the 1980s, several prominent ecological studies to this need beginning in the 1980s, and since then the ability to collect multi-scale observations has grown rapidly. To assess modern ecology's progress in addressing scale, we analyzed the resolution, extent, interval, and duration of observations in 348 observational studies published between 2004-2014. We found that most observations are collected with conventional field techniques, resulting in generally narrow observational scale domains. In the spatial domain, most observations have resolutions of $\leq$1 m$^2$ and extents of $\leq$10,000 ha. In the temporal domain, most observations were either unreplicated or of low frequency ($\geq$1 month interval), and had relatively short durations ($\leq$1 year). Compared to studies conducted prior to the early 2000s, observational durations and resolutions remain largely unchanged, but intervals have become finer and extents larger. Despite such gains, a large gulf exists between the scales at which phenomena are actually observed, and the scales those observations ostensibly represent, revealing portions of space and time not truly captured by replicates, which raises concerns about observational comprehensiveness. Adding to these concerns, scales were not clearly reported in most studies, suggesting that scale is a minor consideration. Journals can help mitigate this problem by implementing scale reporting standards, which can spur ecologists to more rapidly adopt new observational technologies, and thereby close key gaps in current observational domains.}

@@

Our study focused on two dimensions of spatial scale, resolution (grain) and extent, and two of temporal scale, interval and duration (Table 1, and see Methods for full definitions). We also assessed observational scales within two additional dimensions, \emph{actual} extent (the summed area of spatial replicates) and \emph{actual} duration (the summed observational time of temporal replicates). We evaluated these additional dimensions to gain insight into how much the actual scales of observation (i.e. how much space and time is covered by the measurement) differ from the scales that the observations ostensibly represent. This difference may contain important information about how effectively ecological observations characterize ecological phenomena. First, an increasing gap between actual and ostensible observational scale implies greater interpolation or extrapolation of observed measurements, raising the odds of over-leveraging data. Second (and relatedly), since natural systems are frequently complex, non-linear, and non-random \cite{levin_ecosystems_1998,pringle_spatial_2017,rietkerk_regular_2008}, a larger gap may increase the likelihood of encountering unanticipated data challenges such as censoring (\emph{sensu} \cite{efron_efficiency_1977}), as phenomena may resolve themselves in the space or time between observations.

@@

We undertook three analyses to characterize the scale domains of modern ecological observations. We first evaluated the distribution of observations (domain) within each of the four primary dimensions. Next, we examined observational domains within four different two-dimensional juxtapostions (interval versus resolution; interval versus duration; extent versus resolution; extent versus duration). Lastly, we quantified the differences between actual and ostensible scales of observation (i.e. between actual extent and extent, and actual duration and duration), and examined how these differences varied within the domain of the corresponding ostensible scale.

\vspace{10pt}
\noindent \textbf{Results}
\vspace{5pt}
\\
We reviewed 348 papers randomly selected from 42,918 titles published between 2004-2014 in the top 30 (based on 2012 impact factor) ecology-themed journals. We extracted and analyzed scale data from 378 observations of ``natural'' (i.e. non-experimentally manipulated) ecological features that were reported within 133 of the reviewed papers (plus an additional 62 that these cited as the source of observations). The majority of sampled observations were collected using manual field methods (80\%), followed by automated \emph{in situ} sensing techniques (12.4\%), remote sensing (6.9\%), and paleo-reconstruction ($<$0.8\%).

\vspace{5pt}
\noindent \textbf{\emph{Distributions within the four primary dimensions}}\\
In terms of resolution, spatial replicates for the majority (67\%) of observations (across all methods) were $<$1 m$^2$, a further 24\% were 1 m$^2$ up to 1 ha, and the remaining 9\% were $\geq$1 ha (Fig. 1A). These distributions primarily reflect those of field observations, the dominant observational methodology. Examining the distributions for each observational method (Supplementary Figure 1) shows that automated sensing and paleo-reconstruction had resolutions that were generally finer (85\% or more $<$0.1 m$^2$) than field observations (47\% $<$0.1 m$^2$), while the majority of remote observations were much coarser (70\% $\geq$100 m$^2$).

The extent of 19\% of observations was $<$10 ha, 23\% covered 10-1,000 ha, 11\% 1,000-10,000 ha, 19\% 10,000-100,000 ha, 12\% 100,000-1,000,000 ha, and 15\% $>$1,000,000 ha (Fig. 1B).  As with resolution, the extent covered by automated sensing methods tended to be smaller (52\% $<$100 ha) than those of field observations (31\% $<$100 ha), while all but 4\% of remote observations covered areas $\geq$10,000 ha (as did the small number of paleo-reconstructions).

@@

In the temporal dimensions, 37\% of observations were not repeated (Fig. 1C), 17\% were repeated at short intervals (sub-second to daily), 20\% at daily to monthly intervals, 18\% at monthly to yearly intervals, 6\% at yearly to decadal intervals, and 2\% at decadal or greater intervals. With respect to temporally replicated observations (Supplementary Figure 1), automated sensing techniques had the finest intervals (61\% $\leq$1 day; 100\% $\leq$1 year), followed by remote observation (37\% $\leq$1 day; 78\% $\leq$1 year), field observations (17\% $\leq$1 day; 86\% $\leq$1 year), and paleo-reconstructions (21\% $\leq$1 decade).

Duration was one day or less for 31\% of sampled observations (due to lack of temporal replication), while 10\% covered one day to one month, 23\% lasted one month to one year, 27\% covered 1-10 years, and 9\% spanned a decade or more (including several paleoecological studies covering centuries to millennia; Fig. 1D). Paleo-reconstructions naturally had the longest duration (67\% $>$ 1 decade), while just $\sim$40\% of field, automated, and remote observations had durations of 1 year or longer.

\vspace{5pt}
\noindent \textbf{\emph{Spatial and temporal domains}}\\
Juxtaposing these observational dimensions provides further insight into the spatial and temporal domains of observations (Fig. 2). Contrasting resolution with interval reveals that the majority of temporally replicated observations (unrepeated observations were excluded because they lack interval values) had resolutions of 10 cm$^2$-1 m$^2$ and were revisited at daily to yearly intervals (Fig. 2A). A less dense, oblong concentration of observations bounded on the upper left by monthly to yearly observations at 100 m$^2$ resolution and on the lower right by near-daily to monthly observations with 1-10 ha resolution is also evident. The four observational methods occupied substantially different portions of the domain space, as indicated by the locations of their median values (and illustrated further in Supplementary Figure 2): the median domain of field observations was between 0.1-1 m$^2$ of resolution with a monthly interval, whereas remote observations had a coarser median resolution (1,000 m$^2$) but finer median interval ($\sim$1 day). Paleo-reconstructions and automated sensing techniques were both finely resolved (median between 10 cm$^2$-0.01 m$^2$), but automated approaches had an hourly-daily median interval compared to multi-decadal for paleo-reconstructions.

Comparing the interval and duration of temporally replicated observations showed most observations had daily to decadal intervals and durations of $\geq$1 month up to 1 decade (Fig. 2B). The orientation of this concentration shows that interval increases with duration; observations lasting one month to one year tend to have daily to monthly intervals, while those lasting 1 year to 1 decade tend to have yearly to decadal intervals. This tendency is reflected in the median domain locations of the primary observational methods: automated sensing had the finest median interval (hour-day) and shortest duration (month-year), followed by remote sensing (median interval slightly greater than one day and median duration of 1 year), field observations (median monthly interval and duration just over 1 year), and finally paleo-reconstructions (median interval 1 decade and millennial duration).

Contrasting the two spatial dimensions against one another (for all observations) shows a primary concentration of observations of 10 cm$^2$ to nearly 100 m$^2$ resolution with extents ranging from $\sim$1,000 to nearly 1,000,000 ha (Fig. 2C). The second-most prominent concentration consists of higher resolution (1 cm$^2$-1 m$^2$), smaller extent (10-1,000 ha) observations, beneath which lies a third and fainter concentration of 1-1,000 cm$^2$ resolution and 1000 m$^2$ to $<$10 ha extent. These three concentrations suggest a tendency for observational extent to increase with resolution, which is further evident in the median domain values (and kernel densities; Supplementary Figure 2) of automated (0.01 m$^2$ resolution, 100 ha extent), field (0.1-1 m$^2$ resolution, 1,000-10,000 ha extent), and remote (1,000 m$^2$ resolution, 1-10 million ha extent) observations.  Paleo-reconstructions were an outlier from this relationship, having very fine median resolution (0.01 m$^2$) but large extent (1 million ha), a result that likely reflects the very low sample size for this observational type.

Two primary domains of observational concentration are revealed by juxtaposing duration and extent (across all observations). The first consists of observations spanning just over one month to one decade in time and 10-1,000 ha in space, while the second is defined by observations of one year to several decades that cover 10,000 to 1,000,000 ha (Fig. 2D).  Three other notable, but lesser areas of concentration are also evident, including small area observations (0.1-1 ha) covering one month to decade, and short duration, temporally unreplicated observations ($<$1 day) of either 1-10 ha or 10,000-1,000,000 ha.  The median observation from automated sensing (1 year duration, 100 ha extent) lies near the center of the first major concentration, while the median extents of field (1-10,000 ha) and remote (1-10 million ha) observations bound the second major concentration at its upper and lower extents, with the median duration of both observational types falling between 1 month to 1 year.

\vspace{5pt}
\noindent \textbf{\emph{Differences between actual and ostensible scales}}\\
For our final analysis of observational domains, we evaluated the differences between the scales represented by the extent and duration dimensions and the scales that ecological observations actually cover. To make this assessment, we log$_{10}$ transformed and then subtracted the values of i) actual extent from extent and ii) actual duration from duration, yielding the magnitude of difference between each pair of dimensions for each observation. We plotted these values (summarized in box plots) against their corresponding extent/duration values to evaluate whether these differences varied with scale (Fig. 3). Extent was on average 5.6 orders of magnitude larger than actual extent, and this difference increased with extent, reaching a maximum of 8.3 orders of magnitude between 100 million and 1 billion ha of extent (Fig. 3A). The difference fell to 3 orders of magnitude between 1-10 billion ha, a domain that was covered by the $<$2\% of observations that were primarily collected with remote sensing. Remote observations had the smallest mean difference magnitude (1.9), compared to 5.7 or larger for the other three methods (see Supplementary Figure 3 for method-specific plots).

The magnitude of difference between the duration and actual duration of observations was somewhat smaller, averaging 3.4 across all observations, ranging from just under 2 for the shortest durations (hour-day) to over 4 for observations lasting 1 decade to 1 century (Fig. 3B). As with extent, the difference fell substantially for the longest durations (century to 10,000 years), as these domains were covered by paleo-reconstructions (Supplementary Figure 3 in SI), which have effectively no difference between duration and actual duration because coring techniques capture continuous temporal records. The mean difference magnitudes for the other three observing methods ranged from just over 3 (field observations and automated sensing) to nearly 6 (remote observations).

\vspace{5pt}
\noindent \textbf{\emph{Potential biases and uncertainties in quantifying scales}}\\
There were several potential methodological aspects that could have influenced our assessment of ecology's spatial and temporal domains. The first stems from our finding that many studies did not precisely report observational scales, which meant that we had to estimate, rather than simply record, these values for most observations (specifically, in 63\%, 60\%, and 69\% of cases for resolution, extent, and actual extent, and 36\%, 64\%, and 83\% of cases for interval, duration, and effective duration, respectively). The inevitable estimation errors may have biased our findings. However, we attempted to quantify and account for this error by assessing between-observer variability and incorporating this uncertainty into our resampling methodology (Supplementary Results). The resulting confidence intervals (Fig. 1) suggest that it was unlikely that estimation errors unduly influenced our findings.

Another potential source of bias lies within our scale-estimation protocols, chiefly with respect to our rule for estimating resolution (the smallest areal unit of \emph{complete} measurement). We selected this definition for the sake of consistency, but some papers reported resolution as a larger area in which sub-samples were taken. For these, our estimates are finer than what the studies' authors apparently considered to be the resolution of spatial replicates. Our domain estimates would presumably be somewhat different if we had included experimentally manipulated observations. For example, average resolution and duration would likely be finer \cite{tilman_ecological_1989,kareiva_spatial_1988}. Additionally, the token 1 second (Supplemental Results) we used to represent the duration of remotely sensed temporal replicates (which are effectively instantaneous) mean that we underestimated actual duration-duration differences for this observational method (Supplementary Figure 3). However, the relatively low proportion of remote observations suggests that the impact on our overall findings was negligible (Fig. 3).

It is also possible that our findings misrepresent the characteristic scales of ecological observations, due to sampling error. Although we used a randomized title draw to obtain a representative sample, we reviewed just 0.8\% of papers published during the 10 year period. Our sample may therefore under- or over-represent observational coverage in certain scale domains, particularly for analyses broken down by observational method. The case where this is of most concern is for paleo-reconstructions, the small sample size of which resulted in a likely over-estimate of the typical extent of such observations (e.g. Fig. 2B, C; however, the interval and duration values are probably more representative).

Finally, because our review did not include papers beyond 2014, the omission of studies from the most recent years could have introduced bias into our domain estimates. To evaluate this possibility, we used linear regressions to examine whether the relative frequency of observing methods changed during the 10 year study period, and whether our findings regarding scale domains might be different if we included more recent studies. Although our sample size was too small to assign statistical significance, we found a possible positive trend in the use of remote observations and a corresponding decline in field observations over the study period. If these trends were not spurious, they suggest that including studies from 2015-2017 would result in a somewhat larger relative sample of remote observations, which would in turn slightly increase the mean extent of ecological observations (see SI for details of trend analyses).

\vspace{10pt}
\noindent \textbf{Discussion}
\vspace{5pt}
\\
\noindent \textbf{\emph{Insights into the scale domains of modern ecology}}\\
Our results suggest that the scale domains of modern ecological observations are fairly narrow, and are collected primarily with conventional field techniques. Spatially, the majority of observations have grains of $\leq$1 m$^2$ and extents of $\leq$10,000 ha (Fig 1A;B). In the temporal domains, most observations are either un-replicated snapshots, or of low frequency ($\geq$1 month interval; Fig. 1C) and relatively short duration ($\leq$1 year; Fig. 1D).

Contrasting observational dimensions reveals that larger extents are associated with larger spatial replicates (Fig. 2C), while longer durations tend to be associated with longer intervals (Fig. 2B). The latter association primarily reflects a cost-imposed tradeoff between sampling frequency and temporal duration that is characteristic of traditional field-based observation, the dominant observational mode, but is also reflected by the relative domain locations of the four main observational methods, suggesting that high observational frequency poses a durational cost across observational methods. A similar cost tradeoff is also evident in the inverse relationship between resolution and interval that dominates that domain space (Fig. 2A), which again primarily relates to characteristics of field-based observations, in which larger spatial replicates demand greater effort that in turn reduces sampling frequency \cite{kareiva_spatial_1988}. Less obvious is the opposite tradeoff that occurs with remote observation (Supplementary Figure 2), where finer resolution is desired to increase detail, but typically comes at the cost of longer intervals \cite{estes_platform_2016}.

As a result of these tradeoffs, there are several notable gaps in the spatio-temporal domains of ecological observations, which lie in the domains defined by high frequency (daily to sub-daily intervals) observations having 1) high to moderate resolutions ($\geq$1 m$^2$ up to 100 ha; Fig. 2A) and 2) decadal or longer durations (Fig. 2B).  Another hole is evident in the high to moderate resolution, large extent (1 million-10 billion ha) domain (Fig. 2C).

>>>>

\author
{Lyndon Estes$^{\ast1, 2}$, Paul R. Elsen$^{3, 4}$, Tim Treuer$^{4}$, Labeeb Ahmed$^{5}$, \\
Kelly Caylor$^{6, 7}$, Jason Chang$^{5}$, Jonathan J. Choi$^{8}$, and Erle Ellis$^{5}$ \\
\\
\normalsize{$^{1}$Graduate School of Geography, Clark University, Worcester MA 01610, USA}\\
\normalsize{$^{2}$Woodrow Wilson School, Princeton University, Princeton, NJ 08544, USA}\\
\normalsize{$^{3}$Department of Environmental Science, Policy, and Management,}\\
\normalsize{University of California Berkeley, Berkeley, CA 94720, USA}\\
\normalsize{$^{4}$Ecology and Evolutionary Biology, Princeton University, Princeton, NJ 08544, USA}\\
\normalsize{$^{5}$Geography and Environmental Systems, University of Maryland Baltimore County,}\\ 
\normalsize{Baltimore, MD 21250, USA}\\
\normalsize{$^{6}$Department of Geography, University of California Santa Barbara, Santa Barbara, CA 93106, USA}\\
\normalsize{$^{7}$Bren School of Environmental Science and Management,}\\
\normalsize{University of California Santa Barbara, Santa Barbara, CA 93106, USA}\\
\normalsize{$^{8}$Nicholas School of the Environment, Duke University, Durham, NC 27708, USA}\\
\normalsize{$^\ast$To whom correspondence should be addressed; E-mail:  LEstes@clarku.edu.}}
% Include the date command, but leave its argument blank.

@@

\section*{Abstract}
\textbf{To understand ecological phenomena, it is necessary to observe their behavior across multiple spatial and temporal scales. Since this need was first highlighted in the 1980s, technology has opened previously inaccessible scales to observation. To help determine whether there have been corresponding changes in the scales observed by modern ecologists, we analyzed the resolution, extent, interval, and duration of observations (excluding experiments) in 348 studies published between 2004-2014. We found that observational scales were generally narrow because ecologists still primarily use conventional field techniques. In the spatial domain, most observations had resolutions $\leq$1 m$^2$ and extents $\leq$10,000 ha. In the temporal domain, most observations were either unreplicated or infrequently repeated ($>$1 month interval), and $\leq$1 year in duration. Compared to studies conducted prior to 2004, observational durations and resolutions appear largely unchanged, but intervals have become finer and extents larger. We also found a large gulf between the scales at which phenomena are actually observed, and the scales those observations ostensibly represent, raising concerns about observational comprehensiveness. Furthermore, most studies did not clearly report scale, suggesting that it remains a minor concern. Ecologists can better understand the scales represented by observations by incorporating autocorrelation measures, while journals can promote attentiveness to scale by implementing scale reporting standards.}

@@

Our study focused on two dimensions of spatial scale, resolution (grain) and extent, and two of temporal scale, interval and duration (Table 1). We analyzed the observational domains within each of these four dimensions and between pairs of these dimensions. We also assessed two additional dimensions, \emph{actual} extent (the summed area of spatial replicates) and \emph{actual} duration (the summed observational time of temporal replicates), which we used to evaluate how much the actual scales of observation (i.e. how much space and time is covered by the measurement) differ from the scales that they ostensibly represent. These differences may impact how effectively observations characterize ecological phenomena. For one, an increasing gap between actual and ostensible observational scales implies greater interpolation or extrapolation of observed measurements, raising the odds of over-leveraging data. Furthermore, since natural systems are frequently complex, non-linear, and non-random \cite{levin_ecosystems_1998,pringle_spatial_2017,rietkerk_regular_2008}, a larger gap increases the likelihood of data challenges such as censoring (\emph{sensu} \cite{efron_efficiency_1977}), as phenomena may resolve themselves in the space or time between replicates.

@@

\vspace{10pt}
\noindent \textbf{Results}
\vspace{5pt}
\\
We reviewed 348 papers randomly selected from 42,918 titles published between 2004-2014 in the top 30 ecology-themed journals. We extracted scale data from 378 observations of ``natural'' (i.e. non-experimentally manipulated) ecological features reported within 133 of the reviewed papers (plus an additional 62 cited as the source of observations). The majority of sampled observations were collected using conventional field methods (80\%), followed by automated \emph{in situ} sensing techniques (12.4\%), remote sensing (6.9\%), and paleo-reconstruction ($<$0.8\%).

\vspace{5pt}
\noindent \textbf{\emph{Observational domains within individual dimensions}}\\
In terms of resolution, spatial replicates for the majority (67\%) of observations (across all methods) were $\leq$1 m$^2$, a further 24\% were 1 m$^2$ up to 1 ha, and the remaining 9\% were $>$1 ha (Fig. 1A). These distributions primarily reflect those of field observations, the dominant observational methodology. Examining the distributions for each observational method (Supplementary Figure 1) shows that automated sensing and paleo-reconstruction had resolutions that were generally finer (85\% or more $\leq$0.1 m$^2$) than field observations (47\% $\leq$0.1 m$^2$), while the majority of remote observations were much coarser (70\% $>$100 m$^2$).

The extent of 19\% of observations was $\leq$10 ha, 23\% covered 10-1,000 ha, 11\% 1,000-10,000 ha, 19\% 10,000-100,000 ha, 12\% 100,000-1,000,000 ha, and 15\% $>$1,000,000 ha (Fig. 1B).  As with resolution, the extent covered by automated sensing methods tended to be smaller (52\% $\leq$100 ha) than those of field observations (31\% $<$100 ha), while all but 4\% of remote observations covered areas $>$10,000 ha (as did the small number of paleo-reconstructions).

@@

In the temporal dimensions, 37\% of observations were not repeated (Fig. 1C), 17\% were repeated at short intervals (sub-second to daily), 20\% at daily to monthly intervals, 18\% at monthly to yearly intervals, 6\% at yearly to decadal intervals, and 2\% at decadal or greater intervals. Among temporally replicated observations (Supplementary Figure 1), automated sensing had the finest intervals (61\% $\leq$1 day; 100\% $\leq$1 year), followed by remote observation (37\% $\leq$1 day; 78\% $\leq$1 year), field methods (17\% $\leq$1 day; 86\% $\leq$1 year), and paleo-reconstructions (21\% $\leq$1 decade).

Duration was $\leq$1 day for 31\% of sampled observations (due to lack of temporal replication), while 10\% covered one day to one month, 23\% one month to one year, 27\% 1-10 years, and 9\% a decade or more (several paleoecological studies covered centuries; Fig. 1D). Paleo-reconstructions naturally had the longest duration (67\% $>$ 1 decade), while just $\sim$40\% of field, automated, and remote observations had durations exceeding one year.

\vspace{5pt}
\noindent \textbf{\emph{Observational domains within two-dimensions}}\\
Contrasting resolution with interval reveals that the majority of temporally replicated observations (unrepeated observations were excluded because they lack intervals) had resolutions of 10 cm$^2$-1 m$^2$ and were revisited at daily to yearly intervals (Fig. 2A). A less dense, oblong concentration of observations bounded on the upper left by monthly to yearly observations at 100 m$^2$ resolution and on the lower right by near-daily to monthly observations with 1-10 ha resolution is also evident. The four observational methods occupied substantially different domains, as indicated by the locations of their median values (and see Supplementary Figure 2): the median domain of field observations had 0.1-1 m$^2$ resolution and a monthly interval, whereas remote observations had a coarser median resolution (1,000 m$^2$) but finer median interval ($\sim$1 day). Paleo-reconstructions and automated sensing techniques were both finely resolved (median between 10 cm$^2$-0.01 m$^2$), but automated approaches had an hourly-daily median interval compared to multi-decadal for paleo-reconstructions.

Comparing the interval and duration of temporally replicated observations showed most observations had daily to decadal intervals and durations of 1 month to 1 decade (Fig. 2B). The orientation of this concentration shows that interval increases with duration; observations lasting one month to one year tend to have daily to monthly intervals, while those lasting 1 year to 1 decade tend to have yearly to decadal intervals. This tendency is reflected in the domain medians of the primary observational methods: automated sensing had the finest median interval (hour-day) and shortest duration (month-year), followed by remote sensing (median interval slightly greater than one day and median duration of 1 year), field observations (median monthly interval and duration just over 1 year), and finally paleo-reconstructions (median interval 1 decade and millennial duration).

Contrasting the two spatial dimensions against one another shows a primary concentration of observations of 10 cm$^2$ to nearly 100 m$^2$ resolution with extents ranging between $\sim$1,000-1,000,000 ha (Fig. 2C). The second-most prominent concentration consists of higher resolution (1 cm$^2$-1 m$^2$), smaller extent (10-1,000 ha) observations, beneath which lies a third and fainter concentration of 1-1,000 cm$^2$ resolution and 1000 m$^2$ to $\leq$10 ha extent. These three concentrations suggest that observational extent increases with resolution, which is further evident in the median domain values (and kernel densities; Supplementary Figure 2) of automated (0.01 m$^2$ resolution, 100 ha extent), field (0.1-1 m$^2$ resolution, 1,000-10,000 ha extent), and remote (1,000 m$^2$ resolution, 1-10 million ha extent) observations.  Paleo-reconstructions were an outlier from this relationship, having very fine median resolution (0.01 m$^2$) but large extent (1 million ha), a possible artifact of small sample size.

There are two primary observational domains within the contrast between duration and extent. The first consists of observations lasting one month to one decade with extents of 10-1,000 ha, while the second is defined by observations of one year to several decades that cover 10,000-1,000,000 ha (Fig. 2D).  Three other notable, but lesser concentrations are also evident, including small area observations (0.1-1 ha) covering one month to decade, and short duration, temporally unreplicated observations ($\leq$1 day) of either 1-100 ha or 10,000-1,000,000 ha.  The median observation from automated sensing (1 year duration, 100 ha extent) lies near the center of the first major concentration, while the median extents of field (1,000-10,000 ha) and remote (1-10 million ha) observations bound the second major concentration at its upper and lower extents, with the median duration of both observational types falling between 1 month to 1 year.

\vspace{5pt}
\noindent \textbf{\emph{Differences between actual and ostensible scales}}\\
Observational extent was on average 5.6 orders of magnitude larger than actual extent (Fig. 3A). This difference increased with extent, reaching a maximum of 8.3 between 100 million and 1 billion ha of extent, then falling to 3 orders of magnitude between 1-10 billion ha (these extents were covered by $<$2\% of observations, which were primarily collected with remote sensing). Remote observations had the smallest mean difference magnitude (1.9), compared to 5.7 or larger for the other three methods (Supplementary Figure 3).

The magnitudes of difference between observational duration and actual duration was somewhat smaller, averaging 3.4 and ranging from just under 2 for the shortest durations (hour-day) to over 4 for observations lasting 1 decade to 1 century (Fig. 3B). As with extent, the difference fell substantially for the longest durations (century to millenia), as these domains were covered by paleo-reconstructions (Supplementary Figure 3), which have little difference between actual and ostensible duration because coring techniques capture continuous temporal records. The mean difference magnitudes for the other three observing methods ranged from just over 3 (field and automated sensing) to nearly 6 (remote observations).

\vspace{5pt}
\noindent \textbf{\emph{Potential biases and uncertainties in quantifying scales}}\\
Our results were potentially influenced by several methodological issues. First, most studies did not precisely report observational scales, thus we had to estimate, rather than simply record, scale values for the majority of observations (we estimated 63, 60, 69, 36, 64, and 83\% of resolution, extent, actual extent, interval, duration, and actual duration values, respectively). Estimation errors may therefore have biased our findings. We attempted to quantify and account for this error by assessing between-observer variability and incorporating this uncertainty into our resampling methodology (Supplementary Results). The resulting confidence intervals (Fig. 1) suggest that that estimation errors did not unduly influence our findings.

Our scale-estimation protocols may also have introduced bias, particularly our rule for estimating resolution (the smallest areal unit of \emph{complete} measurement). We selected this definition for the sake of consistency, but some papers reported resolution as a larger area in which sub-samples were taken. For these, our estimates were finer than what the studies' authors considered to be the resolution. Our results would also  be somewhat different if we had included observations from experiments. For example, average resolution and duration would likely be finer \cite{tilman_ecological_1989,kareiva_spatial_1988}. Additionally, the token 1 second (Supplemental Methods) we used to represent the duration of remotely sensed temporal replicates (which are effectively instantaneous) caused us to underestimate the differences between their durations and actual durations (Supplementary Figure 3). However, the relatively small number of remote observations suggests that the impact of this bias on our overall findings was negligible.

It is also possible that our findings misrepresent observational domains because of sampling error. Although we randomized our sample to ensure representativeness, we reviewed just 0.8\% of the papers published during our study period. Our sample may therefore under- or over-represent observational coverage in certain domains, particularly for specific methods. This possibility is greatest for paleo-reconstructions, where the small sample size likely resulted in an overestimate of typical observational extent (e.g. Fig. 2B, C; however, the interval and duration values are probably more representative).

Finally, our omission of papers published after 2014 could also have biased our findings. Although our sample size was too small to assign statistical significance, we found a possible positive trend in the use of remote observations and a corresponding decline in field observations over the course of our study period. If these trends were not spurious, they suggest that including studies from 2015-2017 would result in a somewhat larger relative sample of remote observations, which would in turn slightly increase the mean extent of ecological observations (see Supplementary Results for further details).

\vspace{10pt}
\noindent \textbf{Discussion}
\vspace{5pt}
\\
\noindent \textbf{\emph{Insights into the scale domains of modern ecology}}\\
Our results suggest that modern ecology's observational domains are fairly narrow, and that ecologists still primarily rely on conventional field-based observational techniques. In the spatial dimensions, the majority of observations have grains of $\leq$1 m$^2$ and extents of $\leq$10,000 ha (Fig 1A;B). In the temporal dimensions, most observations are either un-replicated or relatively infrequent ($>$1 month interval; Fig. 1C) and have relatively short duration ($\leq$1 year; Fig. 1D).

Contrasting observational dimensions reveals that larger extents are associated with larger spatial replicates (Fig. 2C), while longer durations are associated with longer intervals (Fig. 2B). The latter association reflects a cost-imposed tradeoff between sampling frequency and temporal duration that is characteristic of field observations, but also appears to affect the other three methods, as evidenced by their relative domain locations. A similar tradeoff can be seen in the inverse relationship between resolution and interval (Fig. 2A), which primarily relates to field observations, where larger spatial replicates demand greater effort, which in turn reduces sampling frequency \cite{kareiva_spatial_1988}. Less obvious is the opposite tradeoff that affects remote observation (Supplementary Figure 2), where finer resolution (which is necessary for detail) typically necessitates longer intervals \cite{estes_platform_2016}.

As a result of these tradeoffs, there are several notable observational gaps, specifically within the domains defined by high frequency (daily to sub-daily intervals) observations having 1) high to moderate resolutions ($>$1 m$^2$ up to 100 ha; Fig. 2A) and 2) decadal or longer durations (Fig. 2B).  Another hole is evident in the high to moderate resolution, large extent (1 million-10 billion ha) domain (Fig. 2C).

Have these domains changed since the seminal papers on scale first appeared in the late 1980s \cite{wiens_spatial_1989, levin_problem_1992, tilman_ecological_1989}? A comprehensive answer would require a repeat similar analysis focused on earlier literature, but the data provided by three prior studies provides partial insight.  The first dataset consists of duration values that Tilman \cite{tilman_ecological_1989} extracted from 623 studies published between 1977-1987 in the journal \emph{Ecology}. The mean duration of the most comparable subset of those values (n=419) was 3.6 years, versus 3.3 years in our sample (or 5.1 when excluding temporally un-replicated observations). The second dataset is found in Kareiva and Anderson \cite{kareiva_spatial_1988}, who assessed the resolutions of 97 community ecology experiments published in \emph{Ecology} between 1980-1986. The average of those (12,657 m$^2$) was substantially smaller than the mean of our sample (1,479,465 m$^2$), but comparing Kareiva and Anderson's \cite{kareiva_spatial_1988} 80th percentile value (197 m$^2$) to ours (115 m$^2$) shows that the majority of contemporary observations are finer-grained than most 1980s-era experiments. Porter et al \cite{porter_wireless_2005} provide the third dataset, which compares the extent and interval of 25 studies published in 2003 and 2004 (also in \emph{Ecology}). The mean interval was 178 days, compared to 684 days in our sample, but the 80th percentile value in our study was 169 days compared to 329 days in theirs. Extent in our sample was substantially larger according to multiple summary statistics, including the mean (368,403 ha versus 114,965,072 ha in our study), median (9 ha versus 5,051 ha), and 95th percentile (136,000 ha versus 46,424,808 ha; this value is smaller than the mean, which is skewed by a small number of very large extent observations).

Although limited due to methodological differences (e.g. a focus on experiments versus un-manipulated systems), these comparisons suggest that the duration and resolution of ecological observations have changed little in the past 30 years, but observational frequency and extent have both increased. A weak positive trend in our own data also suggests that the mean extent of ecological observations is steadily increasing (Supplementary Figure 5), which likely corresponds to increasing use of remote sensing (Supplementary Figure 4).

Despite this apparent increase in observational extent, there remains a large gulf between the areas that ecologists actually observe and the areas their observations are intended to represent (Fig. 3A). A substantial discrepancy also exists between the amount of time spent observing phenomena and the time spans those observations theoretically represent (Fig. 3B). These differences between the actual and ostensible scales of observation have implications for ecological understanding, as the unobserved portions of space and time may contain important patterns and processes that are not captured by replicates, due to phenomenon-dependent factors such as autocorrelation and representativeness of the sampling scheme \cite{underwood_experiments_1997,palmer_scale_1994, cao_comparison_2002, legendre_spatial_1993,collins_method_2000}. Brief, infrequent snapshots, or fine-grained, spatially sparse replicates, may be sufficient to characterize many phenomena (e.g. annual changes in tree cover are well-represented by low frequency satellite imaging \cite{hansen_high-resolution_2013}), but may be inadequate for more dynamic phenomena. For example, wildfire extent and duration can be mapped by daily return satellites \cite{roy_prototyping_2005,jones_fire_2009}, but the instantaneous nature of the imaging means that it cannot be used to observe fire behavior \cite{clements_observing_2007}. To capture such behavior, long periods of continuous observation may be more important for understanding dynamics than frequent repeats.
